06_gam

### 2023-11-21 06_gam.Rmd
Update 06_gam.Rmd

Required Packages
```{r setup}
library(tidyverse)
library(mgcv)
```

@@

ZC_GAM_env_TOL <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope +
                        TOL_125 + TOL_2000 + shipin10km + timeAfterShip,
              data = ZC_filt, family = binomial, method = "REML")
summary(ZC_GAM_env_TOL)
gam.check(ZC_GAM_env_TOL)
```

@@

PM_GAM_env_TOL <- gam(Wpresence>0 ~ dist2slope + ttDepth + depth + mldDepth + ssh_mean +
                        TOL_125 + TOL_2000 + timeAfterShip,
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM_env_TOL)
gam.check(PM_GAM_env_TOL)
```

@@

>>>>

Required Packages
```{r setup}
library(tidyverse)
library(mgcv)
library(car)
```

@@

# Environ + noise
ZC_GAM_env_TOL <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope +
                        TOL_125 + TOL_2000,
                      data = ZC_filt, family = binomial, method = "REML")
summary(ZC_GAM_env_TOL)
gam.check(ZC_GAM_env_TOL)

#Environ + noise + ship (** CANT MODEL SHIP)
ZC_GAM_env_TOL <- gam(Wpresence>0 ~ depth + sst_mean + curl_mean + mldTemp + bathy_slope +
                        TOL_125 + TOL_2000 + shipin10km + timeAfterShip,
              data = ZC_filt, family = binomial, method = "REML")
```

@@

# Environ + noise
PM_GAM_env_TOL <- gam(Wpresence>0 ~ dist2slope + ttDepth + depth + mldDepth + ssh_mean +
                        TOL_125 + TOL_2000,
              data = PM_filt, family = binomial, method = "REML")
summary(PM_GAM_env_TOL)
gam.check(PM_GAM_env_TOL)

# With ships - DO NOT MODEL SHIPS
#PM_GAM_env_TOL <- gam(Wpresence>0 ~ dist2slope + ttDepth + depth + mldDepth + ssh_mean +
#                        TOL_125 + TOL_2000 + timeAfterShip,
#              data = PM_filt, family = binomial, method = "REML")
```

@@

Detecting Multicolinearity with Variance Inflation Factors (VIF) - Cuvier's Only
```{r}
ZC_filt <- allDrifts %>% 
  filter(species %in% c("ZC", NA))

# Environ only
model1LM <- glm(Wpresence ~ depth + mldDepth
               + sst_mean + curl_mean + dist2slope
               + bathy_slope + dist2shore 
               + chlorophyll_mean + ssh_mean 
               + ttDepth + temp400 + sal400,
               data = ZC_filt)

vif(model1LM)

vif_values <- tibble(vif(model1LM)) %>%
  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400")) %>%
  rename(val = "vif(model1LM)")
# Need to remove mldTemp or ttTemp for it to work.

ggplot(vif_values, aes(y = var, x = val)) + 
  geom_col() + 
  geom_vline(xintercept = 5, col = "red") +
  labs(title = "VIF Values by Env. Variable - Cuvier's BW only")
```

Detecting Multicolinearity with Variance Inflation Factors (VIF) - Sperm whale only
**IDENTICAL RESULTS TO ZC ONE***
```{r}
PM_filt <- allDrifts %>% 
  filter(species %in% c("PM", NA))

# Environ only
model1LM <- glm(Wpresence ~ depth + mldDepth
               + sst_mean + curl_mean + dist2slope
               + bathy_slope + dist2shore 
               + chlorophyll_mean + ssh_mean 
               + ttDepth + temp400 + sal400,
               data = PM_filt)

vif(model1LM)

vif_values <- tibble(vif(model1LM)) %>%
  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400")) %>%
  rename(val = "vif(model1LM)")
# Need to remove mldTemp or ttTemp for it to work -- perfectly correlated!

ggplot(vif_values, aes(y = var, x = val)) + 
  geom_col() + 
  geom_vline(xintercept = 5, col = "red") +
  labs(title = "VIF Values by Env. Variable - Sperm whales only")
```
Trying GAM with correlated variables removed
```{r}
# Cuvier's BW
zc.GAM.VIF <- gam(Wpresence>0 ~ ttDepth + ssh_mean + mldDepth + depth + curl_mean +
            chlorophyll_mean + bathy_slope, data = ZC_filt)
summary(zc.GAM.VIF)
gam.check(zc.GAM.VIF)

# Sperm whale
pm.GAM.VIF <- gam(Wpresence>0 ~ ttDepth + ssh_mean + mldDepth + depth + curl_mean +
            chlorophyll_mean + bathy_slope + TOL_125 + TOL_2000, data = PM_filt)
summary(pm.GAM.VIF)
gam.check(pm.GAM.VIF)

```

Load seafloor depth gradient dataset and rerun GAM
```{r}
load(file = "data/allDrifts_grad.rda")

ZC_filt2 <- allDrifts_grad %>% 
  filter(species %in% c("ZC", NA))

# Environ only
model2LM <- glm(Wpresence ~ depth + mldDepth
               + sst_mean + curl_mean + dist2slope
               + bathy_slope + dist2shore 
               + chlorophyll_mean + ssh_mean 
               + ttDepth + temp400 + sal400 +
               magnitude_gradient_mean,
               data = ZC_filt2)

vif(model2LM)

vif_values <- tibble(vif(model2LM)) %>%
  cbind(var = c("depth", "mldDepth", "sst_mean", "curl_mean", "dist2slope", "bathy_slope",
          "dist2shore", "chlorophyll_mean", "ssh_mean", "ttDepth", "temp400", "sal400",
          "magnitude_gradient_mean")) %>%
  rename(val = "vif(model2LM)")
# Need to remove mldTemp or ttTemp for it to work.

ggplot(vif_values, aes(y = var, x = val)) + 
  geom_col() + 
  geom_vline(xintercept = 5, col = "red") +
  labs(title = "VIF Values by Env. Variable - Cuvier's BW only")
```

```{r}
# GAM with seafloor gradient

# Cuvier's
model3GAM <- gam(Wpresence ~ depth + mldDepth
               + curl_mean + bathy_slope
               + chlorophyll_mean + ssh_mean +
               magnitude_gradient_mean,
               data = ZC_filt2)
summary(model3GAM)
# Deviance explained = 0.854%

# Sperm whale
PM_filt2 <- allDrifts_grad %>% 
  filter(species %in% c("PM", NA))

model4GAM <- gam(Wpresence ~ depth + mldDepth
               + curl_mean + bathy_slope
               + chlorophyll_mean + ssh_mean +
               magnitude_gradient_mean,
               data = PM_filt2)
summary(model4GAM)
# Deviance explained = 0.816%
```
